Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-09 WSSSPE3/report/wg_main_funding_spec_expert.tex
Update wg_main_funding_spec_expert.tex

\subsection{Research Programming Expertise -- how to fund it?}
\label{RSE}

@@

Research Software Engineers -- those who contribute to science and scholarship
through computer programming -- are an important part of the team
needed to deliver 21st century research. However, existing academic structures
and systems of funding do not effectively fund and sustain these skills.
The resulting high levels of turnover, and inappropriate incentives,
are a significant contributing factor to low levels of reliability and
readability observed in scientific software.

For effective development of software for advanced research, it is necessary
that researchers can work with scientific programmers who are able to understand
the research domain sufficiently to be able to build meaningful software. This
requires a collaborative approach -- programmers fully engaged in the research
context, co-developing software with domain academics.

@@

One solution to this problem would be a successful solution to the problems of
obtaining academic credit for programming work. If this were solved, traditional
academic funding models and career tracks would properly sustain this
contribution to research. However, in the absence of guaranteed success in that
stream, this working party seeks to look to an alternate solution; the
development and funding of dedicated research programming roles. We also noted
that measures of impact and success for research programming
groups, as well as for individual research software engineers, will be required
in order to make the case for continued funding. Research programmers will not
be measured by publications, we hope, but by other measures. Middle-author
publications are common for RSEs. Most RSEs welcome co-authorship on papers
where the PIs consider the contribution deserves it.

A contributing factor to the problem is the devastating effect on an academic
publication record resulting from time in industry; such postings often
develop exactly the skills that research programmers need, yet returns to
university positions following an industry role are penalized by the current
structures. Retention of senior programmers is hard, because these people
are highly in demand by the economy. However,
people who have a PhD in science and enter industry,
may desire to return for diverse reasons, and should be welcomed back.

@@

One solution is to fund dedicated programming roles for major research software
projects at national laboratories or other non-educational institutions. This
solution is working well for many well-used scientific codebases.
However, much of the body of software for which this expertise is needed sits
within the longer tail of research in universities. Therefore, we argue that
institutions need to develop hybrid academic-technical roles for this capacity
who work with more than one PI, rather than the traditional RA role within a
single group. This could be centrally, perhaps within research computing
organizations which have traditionally supported university cyberinfrastructure,
library organizations, or research offices. Alternatively, these groups could
be organizationally closer to research groups, sitting within academic
departments. The most
effective model will vary from institution to institution,
but the mandate and ways of working should be similar.

Having convinced ourselves that this would be a positive innovation, we were
then faced with the specific question of how to fund the initiation of this
activity. A self-sustaining research software group will support itself
through collaborations with PIs in the normal grant process, with PIs choosing
to fund some amount of research programming effort through grants in the usual
way. However, to bootstrap such a function to a level where it has the
reputation and client base to be self-sustaining requires seed investment.

@@

Success, appropriately measured, will help make the case for further investment.
However, measurement of code cleanliness, and the resulting productivity gains,
is an unsolved problem in empirical software engineering. To measure
``what did not go wrong'' because of an intervention is particularly hard.

We finally noted that the institutional case for such groups is made easier
by having successful examples to point to. In the UK, a collective effort to
identify this community, with individuals clearly stating ``I'm a research
software engineer'', has been important to the campaign. It will be useful to
the global effort to similarly identify emerging research software organizations,
and also, importantly, to identify longer-running research software groups,
which have in some cases had a long running \emph{sui-generis} existence, but which
now can be identified as part of a wider solution.

@@

We will therefore seek to gather a list of groups which are prepared to
self-identify as research software engineering groups. We will collect information
as to the organizational models under which these groups function, and how they
are funded.

>>>>

\subsection{Funding Research Programmer Expertise}
\label{RSE}

@@

Research Software Engineers -- those who contribute to science and scholarship
through software development -- are an important part of the team
needed to deliver 21st century research. However, existing academic structures
and systems of funding do not effectively fund and sustain these skills.
The resulting high levels of turnover and inappropriate incentives
are a significant contributing factor to low levels of reliability and
readability observed in scientific software. Moreover, the absence of skilled and experienced 
developers retards progress in key projects, and at times causes important projects to fail completely.

Effective development of software for advanced research requires
that researchers work closely with scientific software developers who understand
the research domain sufficiently to build meaningful software at a reasonable pace. 
This requires a collaborative approach -- where developers who are fully engaged/invested 
in the research context are co-developing software with domain academics.

@@

The solution we envision entails creating an environment where software developers 
are a stable part of the research team. Such an environment mitigates the risk
of losing a key developer at a critical moment in a projects lifetime, and provides
the benefits of building a store of institutional knowledge about specific projects as well as about software 
development for today's research. Our vision is to find a way to promote a University/research institute environment where
software developers are stable components of research project teams.

One strategy to promote stability is implementing a mechansim for developers to obtain academic
credit for software development work. With such a mechanism in place, traditional academic funding
models and career tracks could properly sustain individuals for whom software development is their
primary contribution to research. A contributing factor to the problem with the current academic reward system is the
devastating effect on an academic
publication record resulting from time in industry; such postings often develop exactly the skills that research programmers
need, yet returns to university positions following an industry role are penalized by the current structures. Retention of
senior programmers is hard, because these people are highly in demand by the economy. However, people who have a PhD in
science and enter industry, may desire to return for diverse reasons, and should be welcomed back.

While development of new mechanisms in the current academic reward system is a worthy aspirational goal, such a dramatic
change in this structure does not seem likely in a time scale relevant to this working group. Accordingly, our working party
sought alternate solutions that may be achievable within the context of existing academic structures. The group felt that
developing dedicated research programming roles within the University, and finding stable funding for those individuals is
the most promising mechanism for creating a stable software development staff.

Measures of impact and success for research programming groups, as well as for individual research software engineers, will
be required in order to make the case to the University for continued funding. Research programmers will not be measured by publications, we hope, but by other measures. (what measures?) Middle-author publications are common for RSEs. Most RSEs welcome co-authorship on papers where the PIs consider the contribution deserves it.
(these last two statements seem contradictory)

@@

One solution is to fund dedicated programming roles for major research software projects at national laboratories
or other non-educational institutions. This solution is in place and working well for many well-used scientific codebases.
However, this strategy has limited application, as much of the body of software is created and maintained in research
universities. Therefore, we argue that research institutions should develop hybrid academic-technical tracks for this
capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group
. This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
traditionally supported university cyberinfrastructure, libraryorganizations, or research offices. Alternatively, these
groups could be organizationally closer to research groups, sitting within academic departments. The most effective model
will vary from institution to institution, but the mandate and ways of working should be similar.

Having convinced ourselves that this would be a positive innovation, we were then faced with the specific question of how to
fund the initiation of this activity. A self-sustaining research software group will support itself through collaborations
with PIs in the normal grant process, with PIs choosing to fund some amount of research programming effort through grants in
the usual way. However, to bootstrap such a function to a level where it has sufficent reputation and client base to be self
-sustaining will generally require seed investment.

@@

Success, appropriately measured, will help make the case to such funding bodies for further investment. One might expect that metrics such as improved productivity, software adoption rates, and grant success rates would be sufficient arguments in favor of such a model. However, useful measurement of code cleanliness, and the resulting productivity gains, is an unsolved problem in empirical software engineering. To measure ``what did not go wrong'' because of an intervention is particularly hard.

We finally noted that the institutional case for such groups is made easier by having successful examples to point to. In the
UK, a collective effort to identify this community, with individuals clearly stating ``I'm a research software engineer'',
has been important to the campaign. It will be useful to the global effort to similarly identify emerging research software
organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a
long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution. There remains the problem of how to "sell" the value of this investment to investigators in the university will be problem left to the individual universities, once the initial investment is made. This is an issue best addressed by th eindividual organizations that embark on the plan.

@@

The first step in moving this strategy forward is to gather a list of groups that self-identify as research software engineering groups, and to reach out to other organizations to see if there may be a widespread community of RSE's who do not identify themselves as such at this time. We will collect information as to the organizational models under which these groups function, and how they are funded. For example, how many research universities currently fund people in the RSE track, whether they bear the the RSE moniker or not. Are these developers paid by the University or through a program supported by research grants/individual PIs? How did they bootstrap the developer track to get this started? How successful is the university in getting investigators to pay for fractional RSEs?  We will author a report describing our findings, should funding be available to conduct the investigation.
